Fast Sorting on Modern Computers

### 2005-08-31 docs/tech_report/conclusions.tex
davids email conclusion changes merged in

We found that insertion sort has the best performance of any of the elementary
sorts, with fewer instructions and cache misses than any other elementary sort,
and exactly one branch misprediction per key. Selection sort also has very few
branch mispredictions, and our results match figures created using Knuth's
harmonics formula. We found that bubblesort and shakersort have poor branch
prediction performance, and that this cripples their performance, especially
compared to selection sort which is otherwise similar in behaviour. We
determined that bubblesort has better branch performance at the start and end of
its sorting, then in the middle.

@@

No results for the branch prediction performance of sorting algorithm yet
exists. We discovered some interesting new results.

@@

In heapsort, we found a pattern of branch mispredictions when using a fanout in
our 8-heap, and that this pattern matched closely that of selection sort,
discussed above. We found that though our approximated results using $H_N-1$
were not identical to our actual results - due to the arrays being very small -
our results showed the branches becoming more predictable in a similar fashion.

>>>>

We found that insertion sort has the best performance of any of the elementary
sorts, with fewer instructions and cache misses than any other elementary sort,
and exactly one branch misprediction per key. Selection sort also has very few
branch mispredictions, and our results match figures created using Knuth's
harmonics formula. We found that bubblesort and shakersort have poor branch
prediction performance, and that this cripples their performance, especially
compared to selection sort which is otherwise similar in behaviour. We
determined that bubblesort has better branch performance at the start and end of
its sorting, than in the middle.

@@

To our knowledge, this is the first large-scale, systematic study of branch
prediction in sorting algorithms. We discovered some interesting new results.

@@

In heapsort, we found a pattern of branch mispredictions when using a fanout in
our 8-heap, and that this pattern matched closely that of selection sort,
discussed above. We found that the number of mispredictions was close to, but
measurably higher than, $H_N-1$. The main reason was that the number of items
from which the selection was being made was small. Although the comparisons are
mostly biased in one direction, the early ones are sufficiently unpredictable
that the branch predictor often predicts the less likely direction. The later
comparisons are highly predictable, and yield prediction accuracies much closer
to that indicated by $H_N-1$.
